---
title: "Air Quality Data - PM2.5 FRM/FEM Mass - 88101 "
author: "Fatemeh Kazemi"
date: "02-27-2021"
output:
  html_document:
    df_print: paged
---

### This program:
  (1) Downloads and cleans PM2.5 data measured by EPA
      i. https://aqs.epa.gov/aqsweb/airdata/download_files.html#Daily
  (2) Choose a subset of monitors on the basis of data availability
      i. 4days - 9month- 4years
  (3) Smooths data with 4df per year to fill small gaps
      i.Does not smooth over more than 90 days of missing data
      ii. Divides data into coherent pieces instead
  (4) Calculates yearly average of PM2.5 by site and month
      i. Require at least 350 values to compute yearly average from smoothed data
  (5) Uses cross-walk files between Sites and Zipcodes to assign the measurements
      from Sites to the Zipcodes in their x-mile bufferzone
  (6) Merge it with CMS data (on NEU Cluster) to find the Sites which have
      population living in their x-mile bufferzone
      
### Note:
  1. In data merged with CMS, bufferzone is 6-mile unless pm25ified

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(here)
library(lubridate)
```

```{r Call Functions}
source(here('code','data-processors','epa-air-data-download-function.R'))
source(here('code','data-processors','air-data-processor-function.R'))
source(here('code','data-processors','air-data-smooth-mvavg-function.R'))
```

```{r load data}
dirUSSpatial <- 
  'C:\\Users\\fkazem01\\Box\\Projects\\USA Spatial\\data\\processed\\'
dirpm25 <- 
  'C:\\Users\\fkazem01\\Box\\Projects\\PM Components\\data\\processed\\'
load(paste0(dirUSSpatial,'state-region.RDa'))
load(paste0(dirUSSpatial,'site-zip.RDa'))
load(paste0(dirpm25,'aqs-sites.RDa'))
```

```{r EPA AQS Data Download}
#Using EPA.AQS.download Function
dt.pm25.raw <- EPA.AQS.download (period = "daily", Index = "88101",
                        start = 1999, end = 2008)
dt.pm25.raw$Parameter.Name <- "PM25"
save(dt.pm25.raw, file = here('data', 'raw', 'pm25-raw.RDa'))
#Monitor.ID: Site.ID-POC

monitor.pm <- dt.pm25.raw %>% distinct(Monitor.ID) #1906
site.pm <- dt.pm25.raw %>% distinct(Site.ID)#1540

dt.pm25.raw %>% 
  mutate(Year = year(Date)) %>% 
  distinct(Site.ID, Year) %>% 
  group_by(Year) %>% 
  summarise(n = n())
```

```{r Air quality data Process}
# Find sites with at least 4day, 9month and 4 year
# Using Air.Daily.Process function
dt.pm25.sel <- Air.Daily.Process(dt = dt.pm25.raw, d = 4, m = 9, y = 4)

monitor.pm25 <- dt.pm25.sel %>% distinct(Monitor.ID) # 1906 > 1178
site.pm25 <- dt.pm25.sel %>% distinct(Site.ID) # 1540 > 1039

# POC: Parameter Occurrence Code ...
# ... more than one instrument measured a component at one site
# pick the POC(instrument) with larger number of measurements
site.POC.larger <- dt.pm25.sel %>% 
  group_by(Site.ID, POC) %>% 
  summarise(n = n()) %>% 
  group_by(Site.ID) %>% 
  filter(n == max (n)) %>% 
  select(Site.ID, POC)

dt.pm25.sel <- merge(dt.pm25.sel, site.POC.larger) 
monitor.pm25 <- dt.pm25.sel %>% distinct(Monitor.ID) # 1906 > 1178 > 1039
site.pm25 <- dt.pm25.sel %>% distinct(Site.ID) # 1540 > 1039
dt.pm25.sel$POC <- dt.pm25.sel$Monitor.ID <- NULL
save(dt.pm25.sel, file = here('data', 'processed', 'pm25-selected.RDa'))
```

```{r Smoothing Data & Caluculating 1-yr mvavg for each month}
dt.air <- dt.pm25.sel %>% 
  rename(Parameter.Symbol = Parameter.Name,
         Value = Arithmetic.Mean) %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

site.para <- dt.air %>% 
  distinct(Site.ID, Parameter.Symbol)

#using smoothed.yr function
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 0.99, 350)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.1yr=unlist(smooth.out[4,]))

dt.pm25.1yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.1yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.1yr, file = here('data', 'processed', 'pm25-1yr.RDa'))

dt.site.pm25 <- dt.pm25.1yr %>% distinct(Site.ID) %>% 
  merge(dt.aqs.sites) #1039
save(dt.site.pm25, file = here('data', 'processed', 'site-pm25.RDa'))

dt.site.pm25 %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
#Location.Setting  n
#                  1			
#RURAL	           178			
#SUBURBAN	         441		
#URBAN	           419

dt.pm25.1yr %>% 
  distinct(Site.ID, Year) %>% 
  group_by(Site.ID) %>% 
  summarise(n = n()) %>% filter(n > 4) %>% 
  nrow() #1039: all  #951: with 4+ years

check <- dt.pm25.1yr %>% 
  group_by(Site.ID) %>% 
  summarise(n = n()) %>% 
  arrange(n)

summary(check$n) # 25 to 109
```

```{r Zipcodes in x-mile bufferzone of Monitoring Sites}
# remove components for now to make the dataset smaller
dt.pm25 <- dt.pm25.1yr %>% 
  select(Site.ID, Year, Month) %>% 
  mutate(psuedo = 100)

# joining zipcodes to monitoring sites
dt.pm25 <- dt.pm25 %>% 
  inner_join(dt.site.zip, by = c("Site.ID", "Year")) %>% 
  arrange(Zip.Code, Year, Dist)

# for each zipcode, pick the site with shorter distance to zipcode centroeid
dt.site.dist.shorter <- dt.pm25 %>% 
  group_by(Zip.Code, Year) %>% 
  filter(Dist == min (Dist)) %>% 
  arrange(Zip.Code, Year) %>% 
  as.data.frame()

# 6-mile bufferzone around monitoring site
dt.pm25.site.zip.bz6 <- dt.site.dist.shorter %>% 
  filter(Dist <= 6) %>% 
  select(Zip.Code, Year, Month, Site.ID, psuedo)

save(dt.pm25.site.zip.bz6, 
     file = here('data', 'processed', 'pm25-site-zip-bz6.RDa'))
write.csv(dt.pm25.site.zip.bz6, 
          file = here('data', 'processed', 'pm25-site-zip-bz6.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pm25.site.zip.bz6 %>% distinct(Zip.Code) %>% nrow() #4348
dt.pm25.site.zip.bz6 %>% distinct(Site.ID) %>% nrow() #322 > 274

# 12-mile bufferzone around monitoring site
dt.pm25.site.zip.bz12  <- dt.site.dist.shorter %>% 
  filter(Dist <= 12) %>% 
  select(Zip.Code, Year, Month, Site.ID, psuedo)

save(dt.pm25.site.zip.bz12, 
     file = here('data', 'processed', 'pm25-site-zip-bz12.RDa'))
write.csv(dt.pm25.site.zip.bz12, 
          file = here('data', 'processed', 'pm25-site-zip-bz12.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pm25.site.zip.bz12 %>% distinct(Zip.Code) %>% nrow() #8446
dt.pm25.site.zip.bz12 %>% distinct(Site.ID) %>% nrow() #322 > 320

# 24-mile bufferzone around monitoring site
dt.pm25.site.zip.bz24  <- dt.site.dist.shorter %>% 
  filter(Dist <= 24) %>% 
  select(Zip.Code, Year, Month, Site.ID, psuedo)

save(dt.pm25.site.zip.bz24, 
     file = here('data', 'processed', 'pm25-site-zip-bz24.RDa'))
write.csv(dt.pm25.site.zip.bz24, 
          file = here('data', 'processed', 'pm25-site-zip-bz24.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pm25.site.zip.bz24 %>% distinct(Zip.Code) %>% nrow() #15778
dt.pm25.site.zip.bz24 %>% distinct(Site.ID) %>% nrow() #322 > 321

# pm25-site-zip-bz12.csv was uploaded on NEU Cluster to be merged with CMS data
# It has 320 sites with 8446 zipcodes around them
# 317 out of 320 sites have population living in their 12-mile bufferzone

# pm25-site-zip-bz6.csv was uploaded on NEU Cluster to be merged with CMS data
# 271 out of 274 sites have population living in their 6-mile bufferzone
# We extended the bufferzone around these 271 sites to 12- and 24- miles too
```

```{r Post mergeing with CMS Data}
dt.pm25.site.ndi <- read.csv(here('data', 'processed', 'site_pm25_ndi.csv')) %>% 
  rename(Site.ID = Site_ID) %>% 
  left_join(dt.aqs.sites.ii) %>% 
  left_join(dt.site.carbon.method)
# 317 out of 320 sites have population living in their 12-mile bufferzone

summary(dt.pm25.site.ndi$Location.Setting) 
fix(dt.pm25.site.ndi) # changed blanks to 'RURAL'

save(dt.pm25.site.ndi, file = here('data', 'processed', 'pm25-site-ndi.RDa'))
write.csv(dt.pm25.site.ndi, 
          file = here('data', 'processed', 'pm25-site-ndi.csv'),
          row.names = F)

# TOT/TOR Method in different location.set
dt.pm25.site.ndi %>% 
  group_by(Location.Setting, Method) %>% 
  summarise(n = n())

# urban and suburban: TOT >> TOR
# rural:              TOR >> TOT
# improve sites only in rural

dt.pm25.site.ndi %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
# urban 72
# suburban 69
# rural 176

dt.pm25.site.ndi %>% 
  group_by(Region.IV, Location.Setting) %>% 
  summarise(n = n())
#1	102			
#2	68			
#3	93			
#4	54	


dt.pm25.site.ndi.bz6 <- read.csv(here('data', 'processed', 'site_pm25_ndi_bz6.csv')) %>% 
  rename(Site.ID = Site_ID) %>% 
  left_join(dt.aqs.sites.ii) 

dt.pm25.site.ndi.bz6 %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
```

```{r - daily data for eligible sites}
dt.pm25.daily <- dt.pm25.site.ndi %>% 
  select(Site.ID) %>% 
  merge(dt.pm25.sel)

save(dt.pm25.daily, file = here('data', 'processed', 'pm25-daily.RDa'))
```




```{r Smoothing Data & Caluculating 1 to 5-yr mvavg for each month}
cmp.sub <- c("CA", "CL", "NA", "NI", "NO3", "OC", 
             "PB", "PM", "SE", "SI", "SO4", "V", "ZN")
dt.air <- dt.pm25.daily %>% 
  select(Site.ID, Date, cmp.sub) %>% 
  pivot_longer(cols = cmp.sub,
               names_to = "Parameter.Symbol",
               values_to = "Value") %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

site.para <- dt.air %>% 
  distinct(Site.ID, Parameter.Symbol)

#using smoothed.yr function
# 5 years
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 4.99, 1750)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.yr=unlist(smooth.out[4,]))

dt.pm25.5yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.5yr, file = here('data', 'processed', 'pm25-sub-5yr.RDa'))


# 4 years
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 3.99, 1400)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.yr=unlist(smooth.out[4,]))

dt.pm25.4yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.4yr, file = here('data', 'processed', 'pm25-sub-4yr.RDa'))

dt.pm25.4yr.5yr <- dt.pm25.5yr %>% 
  select(Site.ID, Year, Month) %>% 
  inner_join(dt.pm25.4yr)

save(dt.pm25.4yr.5yr, file = here('data', 'processed', 'pm25-sub-4yr-5yr.RDa'))

# 3 years
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 2.99, 1050)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.yr=unlist(smooth.out[4,]))

dt.pm25.3yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.3yr, file = here('data', 'processed', 'pm25-sub-3yr.RDa'))

dt.pm25.3yr.5yr <- dt.pm25.5yr %>% 
  select(Site.ID, Year, Month) %>% 
  inner_join(dt.pm25.3yr)

save(dt.pm25.3yr.5yr, file = here('data', 'processed', 'pm25-sub-3yr-5yr.RDa'))

# 2 years
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 1.99, 700)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.yr=unlist(smooth.out[4,]))

dt.pm25.2yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.2yr, file = here('data', 'processed', 'pm25-sub-2yr.RDa'))

dt.pm25.2yr.5yr <- dt.pm25.5yr %>% 
  select(Site.ID, Year, Month) %>% 
  inner_join(dt.pm25.2yr)

save(dt.pm25.2yr.5yr, file = here('data', 'processed', 'pm25-sub-2yr-5yr.RDa'))

# 1 year
smooth.out <- mapply(Smoothed.Year, 
                     site.para$Site.ID, site.para$Parameter.Symbol, 0.99, 350)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.yr=unlist(smooth.out[4,]))

dt.pm25.1yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.pm25.1yr, file = here('data', 'processed', 'pm25-sub-1yr.RDa'))

dt.pm25.1yr.5yr <- dt.pm25.5yr %>% 
  select(Site.ID, Year, Month) %>% 
  inner_join(dt.pm25.1yr)

save(dt.pm25.1yr.5yr, file = here('data', 'processed', 'pm25-sub-1yr-5yr.RDa'))

dt.pm25.5yr %>% distinct(Site.ID, Year) %>% nrow() #263 -  820
dt.pm25.4yr %>% distinct(Site.ID, Year) %>% nrow() #303 - 1128
dt.pm25.3yr %>% distinct(Site.ID, Year) %>% nrow() #314 - 1442
dt.pm25.2yr %>% distinct(Site.ID, Year) %>% nrow() #317 - 1745
dt.pm25.1yr %>% distinct(Site.ID, Year) %>% nrow() #317 - 2074
```

```{r}
load(here('data','processed', 'pm25-daily.RDa'))
load(here('data','processed','pm25-site-ndi.RDa'))

dt.pm25 <- dt.pm25.site.ndi %>% 
  mutate(Loc.Set = ifelse(Location.Setting == "RURAL", "Rural", "Urban"),
         Region = ifelse(Region.IV == 1, "W",
                         ifelse(Region.IV == 2, "MW",
                                ifelse(Region.IV == 3, "S", "NE")))) %>% 
    select(Site.ID, Region, Loc.Set) %>% 
  merge(dt.pm25.daily) %>% 
  pivot_longer(AS:PM,
               names_to = "Parameter",
               values_to = "Value")

table.region <- dt.pm25 %>% 
  group_by(Parameter, Region) %>% 
  summarise(Mean = sprintf("%0.2f", mean(Value*1000)),
            STD = sprintf("%0.2f", sd(Value*1000)))

table.urban <- dt.pm25 %>% 
  group_by(Parameter, Loc.Set) %>% 
  summarise(Mean = sprintf("%0.2f", mean(Value*1000)),
            STD = sprintf("%0.2f", sd(Value*1000))) %>% 
  rename(Region = Loc.Set)

table <- dt.pm25 %>% 
  group_by(Parameter) %>% 
  summarise(Mean = sprintf("%0.2f", mean(Value*1000)),
            STD = sprintf("%0.2f", sd(Value*1000))) %>% 
  mutate(Region = "US") %>% 
  as.data.frame() %>% 
  add_row(table.region) %>% 
  add_row(table.urban) %>% 
  pivot_wider(names_from = c(Region),
              values_from = c(Mean, STD))

write.csv(table,
          file = here('output','table', 'pm25-daily-summary.csv'),
          row.names = F)
  
```

